\texttt{Davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-10-27 paper/main.tex
added clarification re: virtual environments & note re: pyproject.toml in section 4

\begin{figure}[tp] \centering
\includegraphics[width=0.6\textwidth]{figs/shareable_code_2d} \caption{\small
\textbf{Systems for ensuring code reproducibility within the Python ecosystem.}
The $x$-axis denotes the ``burden'' placed on users to install and configure
the given system (systems placed further to the right, that fall within the
redder shading, impose a higher setup cost on the user). The $y$-axis denotes
the degree to which the system guarantees that the code will run similarly for
different users (systems placed higher up, that fall within the bluer shading,
offer stronger guarantees). From left to right, bottom to top: plain-text
\textbf{Python scripts} (\texttt{.py} files) provide the most basic ``system''
for sharing raw code. Scripts may reference external packages, but those
packages must be manually installed on other users' systems. Further, any
checking needed to verify that the correct versions of those packages were
installed must also be performed manually. \textbf{Jupyter notebooks}
(\texttt{.ipynb} files) comprise embedded text, executable code, and media
(including rendered figures, code output, etc.). When the \textbf{Davos
package} is imported into a Jupyter notebook, the notebook's functionality is
extended to automatically install any required external packages (at their
correct versions, when specified) into an isolated directory kept separate from
the user's existing set of packages. \textbf{Virtual environments} similarly
allow users to isolate installed packages in a specific directory, and can be
populated with packages from a shared dependency specification
file~\citep[e.g., a requirements.txt or pyproject.toml; ][]{CannEtal16}.
\textbf{Python environments} (again, shared in the form of a dependency
specification file; e.g., an environment.yml) extend this idea to allow
automatically installing a specific version of Python itself. However, both of
virtual environments and Python environments require users to create, populate,
and manage environments manually. This typically entails distributing a
configuration file (e.g., a \texttt{requirements.txt},
\texttt{pyproject.toml}~\cite{CannEtal16}, or \texttt{environment.yml} file)
that specifies all project dependencies (including version numbers) alongside
the primary code base. Users can then install a third-party
tool~\cite[e.g.,][]{BickEtal07, Eust19, Anac12} to read the file and build the
environment. \textbf{Containers} provide a means of defining an isolated
environment that includes a complete operating system (independent of the
user's operating system), in addition to (optionally) specifying a virtual
environment or other configurations needed to provide the necessary computing
environment. Containers are typically defined using specification files (e.g.,
a plain-text \texttt{Dockerfile}~\cite{Merk14}) that instruct the
virtualization engine regarding how to build the containerized environment.
\textbf{Virtual machines} provide a complete hardware-level simulation of the
computing environment. In addition to simulating specific hardware, virtual
machines (typically specified using binary image files) must also define
operating system-level properties of the computing environment. $\Delta_1$
represents the increased reproducibility \texttt{Davos} provides over standard
Jupyter notebooks for no greater setup cost. $\Delta_2$ represents
\texttt{Davos}'s lower setup cost compared to standard virtual environments
despite more stringently ensuring reproducibility.} \label{fig:code-sharing}
\end{figure}

@@

While the \texttt{Davos} package by no means offers a universal solution to this problem, it can, in many cases, provide an effective yet more accessible alternative to existing systems for sharing reproducible Python code, such as specification files for building virtual environments, Python environments, containers, and virtual machines (Fig.~\ref{fig:code-sharing}).
For researchers, this can lower barriers to collaborating with peers and contributing to publicly available open science resources.
And by eliminating most of the setup costs of reconstructing the original researchers' computing environment(s), \texttt{Davos} also lowers barriers to entry for members of the scientific community and the public who seek to run shared code.

Among common systems for sharing reproducible Python code (see Fig.~\ref{fig:code-sharing}), \texttt{Davos} is most comparable to specification files for building virtual environments in that it provides a lightweight mechanism for specifying and sharing a complete set of packages (and specific package versions) required by a particular project, and installing those packages into an isolated directory on the user's file system.
However, when used in conjunction with Jupyter notebooks, \texttt{Davos} offers a number of advantages over standard virtual environments that make it both easier to use and more effective at ensuring reproducibility of shared code.

@@

Finally, neither virtual environments nor \texttt{Davos}-enhanced notebooks can bundle specific versions of Python with which to run shared code.
However, \texttt{Davos}'s \texttt{require\_\-python()} function provides a simple mechanism for indicating a required or constrained Python version and alerting the user at runtime (by raising an exception) if their Python version is incompatible.
In terms of \texttt{Davos}'s ability to ensure that shared code can be executed reproducibly by other users, this falls short of the capabilities of more complex tools that can provide complete copies of Python (Fig.~\ref{fig:code-sharing}).
However, the functionality \texttt{Davos} provides over what is possible with a standard virtual environment is to remove the \textit{expectation} that running a particular notebook will reproduce an expected outcome in situations where this is either impossible or not guaranteed.
With this understanding, a user may choose to install a compatible Python version through some other means or elect to still run the code, but will not be surprised by a potential failure to execute successfully or output an expected result.

>>>>

\begin{figure}[tp] \centering
\includegraphics[width=0.6\textwidth]{figs/shareable_code_2d} \caption{\small
\textbf{Systems for ensuring code reproducibility within the Python ecosystem.}
The $x$-axis denotes the ``burden'' placed on users to install and configure
the given system (systems placed further to the right, that fall within the
redder shading, impose a higher setup cost on the user). The $y$-axis denotes
the degree to which the system guarantees that the code will run similarly for
different users (systems placed higher up, that fall within the bluer shading,
offer stronger guarantees). From left to right, bottom to top: plain-text
\textbf{Python scripts} (\texttt{.py} files) provide the most basic ``system''
for sharing raw code. Scripts may reference external packages, but those
packages must be manually installed on other users' systems. Further, any
checking needed to verify that the correct versions of those packages were
installed must also be performed manually. \textbf{Jupyter notebooks}
(\texttt{.ipynb} files) comprise embedded text, executable code, and media
(including rendered figures, code output, etc.). When the \textbf{Davos
package} is imported into a Jupyter notebook, the notebook's functionality is
extended to automatically install any required external packages (at their
correct versions, when specified) into an isolated directory kept separate from
the user's existing set of packages. \textbf{Virtual environments} similarly
allow users to isolate installed packages in a specific directory, and can be
populated with packages from a shared dependency specification
file (e.g., a \texttt{requirements.txt} or \texttt{pyproject.toml~\cite{CannEtal16})}.
\textbf{Python environments} (again, shared in the form of a dependency
specification file; e.g., an \texttt{environment.yml}) extend this idea to allow
installing a specific version of Python itself. However, both
virtual environments and Python environments require users to create, populate,
and manage environments manually. This typically entails distributing a
configuration file (e.g., a \texttt{requirements.txt},
\texttt{pyproject.toml}~\cite{CannEtal16}, or \texttt{environment.yml} file)
that specifies all project dependencies (including version numbers) alongside
the primary code base. Users can then install a third-party
tool~\cite[e.g.,][]{BickEtal07, Eust19, Anac12} to read the file and build the
environment. \textbf{Containers} provide a means of defining an isolated
environment that includes a complete operating system (independent of the
user's operating system), in addition to (optionally) specifying a virtual
environment or other configurations needed to provide the necessary computing
environment. Containers are typically defined using specification files (e.g.,
a plain-text \texttt{Dockerfile}~\cite{Merk14}) that instruct the
virtualization engine regarding how to build the containerized environment.
\textbf{Virtual machines} provide a complete hardware-level simulation of the
computing environment. In addition to simulating specific hardware, virtual
machines (typically specified using binary image files) must also define
operating system-level properties of the computing environment. $\Delta_1$
represents the increased reproducibility \texttt{Davos} provides over standard
Jupyter notebooks for no greater setup cost. $\Delta_2$ represents
\texttt{Davos}'s lower setup cost compared to standard virtual environments
despite more stringently ensuring reproducibility.} \label{fig:code-sharing}
\end{figure}

@@

While the \texttt{Davos} package by no means offers a universal solution to this problem, it can, in many cases, provide an effective yet more accessible alternative to existing systems for sharing reproducible Python code, such as instructions and specification files for building and populating virtual environments, Python environments, containers, and virtual machines (Fig.~\ref{fig:code-sharing}).
For researchers, this can lower barriers to collaborating with peers and contributing to publicly available open science resources.
And by eliminating most of the setup costs of reconstructing the original researchers' computing environment(s), \texttt{Davos} also lowers barriers to entry for members of the scientific community and the public who seek to run shared code.

Among common systems for sharing reproducible Python code (see Fig.~\ref{fig:code-sharing}), \texttt{Davos} is most comparable to instructions and specification files for building and populating virtual environments in that it provides a lightweight mechanism for specifying and sharing a complete set of packages (and specific package versions) required by a particular project, and installing those packages into an isolated directory on the user's file system.
However, when used in conjunction with Jupyter notebooks, \texttt{Davos} offers a number of advantages over standard virtual environments that make it both easier to use and more effective at ensuring reproducibility of shared code.

@@

Finally, neither specification files used to install packages into virtual environments nor \texttt{smuggle} statements and onion comments included \texttt{Davos}-enhanced notebooks can be used to automatically install a specific version of Python with which to run shared code. 
However, \texttt{Davos}'s \texttt{require\_\-python()} function provides a simple mechanism for indicating a required or constrained Python version and alerting the user at runtime (by raising an exception) if their Python version is incompatible. 
We note that a similar a similar constraint can be imposed by setting the ``\texttt{requires-python}'' attribute in a \texttt{pyproject.toml} file, though creating a virtual environment from a \texttt{pyproject.toml} would require installing an additional third-party tool (such as~\cite{Eust19}).
In terms of \texttt{Davos}'s ability to ensure that shared code can be executed reproducibly by other users, this falls short of the capabilities of more complex tools that can provide complete copies of Python (Fig.~\ref{fig:code-sharing}).
However, the functionality \texttt{Davos} provides over what is possible with a standard virtual environment is to remove the \textit{expectation} that running a particular notebook will reproduce an expected outcome in situations where this is either impossible or not guaranteed.
With this understanding, a user may choose to install a compatible Python version through some other means or elect to still run the code, but will not be surprised by a potential failure to execute successfully or output an expected result.
